Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-21 final_metrics_analysis.Rmd
Interpretations and post-processing stuff

# for Dianne
# use_condaenv("base")

# for Amrita
#use_condaenv("anaconda3")
```

@@

  # row for baseline model
  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
  # row for mitigated bias
  fair_metrics.loc[model] = fair_metrics_mitigated
  metrics_len = len(cols)
  fig, ax = plt.subplots(figsize=(20,4), ncols=metrics_len, nrows=1)
  plt.subplots_adjust(
      left    =  0.125, 
      bottom  =  0.1, 
      right   =  0.9, 
      top     =  0.9, 
      wspace  =  .5, 
      hspace  =  1.1
  )
  y_title_margin = 1.2
  plt.suptitle("Fairness metrics", y = 1.09, fontsize=20)
  sns.set(style="dark")
  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4]
  rect = [-0.1,-0.1,-0.1,0.8]
  bottom = [-1,-1,-1,0]
  top = [1,1,1,2]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
  for i in range(0,metrics_len):
      plt.subplot(1, metrics_len, i+1)
      ax = sns.barplot(x=fair_metrics.index[1:len(fair_metrics)], y=fair_metrics.iloc[1:len(fair_metrics)][cols[i]])

@@

```{python crosstable of recidivism by race}
crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
crosstable
```

@@

```{python exploratory age by race plot}
# age distribution by race - hacking solution for purple dotted line for now
ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
kdeline = ax.lines[0]
mean_black = df_viz.groupby('race').age.median()[0]
xs = kdeline.get_xdata()
ys = kdeline.get_ydata()
height_black = 0.027 # np.interp(mean_black, xs, ys)
ax.vlines(mean_black, 0, height_black, color=purple, ls=':')

mean_white = df_viz.groupby('race').age.median()[1]
height_white = np.interp(mean_white, xs, ys)
ax.vlines(mean_white, 0, height_white, color=green, ls=':')

plt.title('Distribution of Age by Race')
```

@@

```{python exploratory priors by race}
priors_by_race = sns.displot(x="priors_count", row = "race", col = "sex", hue ="race", data=df_viz, palette=workshop_palette)

#plt.title("Prior Charges Count by Race and Sex)
plt.show()
```

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

# for Dianne
use_condaenv("base")

# for Amrita
use_condaenv("anaconda3")
```

@@

  # row for baseline model
  fair_metrics.loc['Baseline'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
  # row for mitigated bias
  fair_metrics.loc[model] = fair_metrics_mitigated
  metrics_len = len(cols)
  fig, ax = plt.subplots(figsize=(20,4), ncols=metrics_len, nrows=1)
  plt.subplots_adjust(
      left    =  0.125, 
      bottom  =  0.1, 
      right   =  0.9, 
      top     =  0.9, 
      wspace  =  .5, 
      hspace  =  1.1
  )
  y_title_margin = 1.2
  plt.suptitle("Fairness metrics", y = 1.09, fontsize=20)
  sns.set(style="dark")
  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4]
  rect = [-0.1,-0.1,-0.1,0.8]
  bottom = [-1,-1,-1,0]
  top = [1,1,1,2]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
  for i in range(0,metrics_len):
      plt.subplot(1, metrics_len, i+1)
      ax = sns.barplot(x=fair_metrics.index[1:len(fair_metrics)], y=fair_metrics.iloc[1:len(fair_metrics)][cols[i]])

@@

```{python crosstable of recitivism by race}
crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
crosstable
```

@@

```{python exploratory age by race plot}
# age distribution by race - hacking solution for purple dotted line for now
#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
#kdeline = ax.lines[0]
#mean_black = df_viz.groupby('race').age.median()[0]
#xs = kdeline.get_xdata()
#ys = kdeline.get_ydata()
#height_black = 0.027 # np.interp(mean_black, xs, ys)
#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')

#mean_white = df_viz.groupby('race').age.median()[1]
#height_white = np.interp(mean_white, xs, ys)
#ax.vlines(mean_white, 0, height_white, color=green, ls=':')

#plt.title('Distribution of Age by Race')
```

@@

```{python exploratory recidivism by race table}
# table of recidivism by race
pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
```

```{python exploratory priors by race}
priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)

plt.title('Prior Charges Count by Race')
plt.show()
```

```{python exploratory priors by sex}
priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)

plt.title('Prior Charges Count by Sex')
plt.show()
```

@@

The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%. The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%. The percentage of false negatives is lower, whereas the true negative is low at 14.33%. This indicates that the model does not do a good job in predicting the defendents who don't recidivate. It is very likely that it will incorrectly predict that someone will recidivate.

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .367 is much higher than 0, the value of fairness.

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .321 is much higher than 0, the value of fairness.

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0.

@@

The value of 3.3 indicates that the model is highly in favor of Black defendents, as it is much greater than 1. This is a hugely different from the baseline model which favors white defendants.

@@

The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%. The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%. The percentage of false negatives is somewhat high for this baseline model, whereas the false positive rate is very low at 2.05%. This indicates that the model does a good job in predicting the defendents who don't recidivate. It is very unlikely that it will incorrectly predict that someone will recidivate.

```{python}
#metrics_per_group(y_test, y_pred_CEO)
#plt.show()
```

@@

At the value of 0.134, the statistical parity difference is now positive and slightly in favor of black defendants.

@@

There has been a significant change from -.408 to 0.088, showing that now the value is within the bounds of "fair" (-0.1 to 0.1)

@@

The baseline model with the average odds difference was -0.438 and now it is 0.15 which indicates being slightly in favor of black defendants.

@@

The baseline value was 0.466 which is highly out of the range of "fair" which goes up to 1. Now at a value of 1.2 it is much higher than 1.
